---
title: "Pop Assignment"
author: "Christen Bossu"
date: "1/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Libraries needed
I'll have you download multiple programs that will not only allow you to assign individuals to breeding group based on their genotypes, but also plot this information in space. 

```{r}
library(raster)  # important to load before tidyverse, otherwise it masks select()
library(tidyverse)
library(sf)
library(ggspatial)
library(rubias)
#install.packages("OriGen", repos="http://R-Forge.R-project.org")
library(OriGen)
library(RColorBrewer)
library(ggplot2)
library(ggpubr)
#This is the function that calculates the distance betwen predicted lat/long and the lat/long in the meta data file
library(geosphere)
library(rworldmap)
library(rgdal)
library(spatialEco)
library(dplyr)
library(reshape2)
library(xlsx)

```


#### White-bellied Pangolin Shapefile

Finally, we need to have a GIS Shapefile that tells us the range of the white-bellied pangolin, so that genoscape can be clipped properly.  We read this shapefile with the 
`st_read()` function from package `sf`. 
See code below. This also comes in handy for looking at probabilty of spatial assignment in Origen, the second analysis we'll do today. You can get this file on the shared Google Drive: BGP_Data_Share. 
```{r}
library(rmapshaper)
breeding <- st_read("data/Phataginus_tricuspis_shapefile/data_0.shp")
```
If you want to quickly plot the shapefile. 

```{r}
plot(breeding)
```

## Population assignment using rubias

Rubias is an R software program that probabilistically assigns individuals to specific reporting units (i.e. genetic clusters diagnosed in the breeding region). Originally used for fish stock identification, we've co-opted it for the pangolin genoscape! I've attached the two rubias files you'll need. The known breeder/georeferenced genotypes and the genotypes of the confiscated pangolin scales. Note, since we don't know where the pangolin scales are coming from, we consider them unknown and we are going to use rubias to estimate mixing proportions and probability of assignment to their breeding origin. Thus, we consider the unknown scales a potential mixture of individuals from distinct genetic clusters.

```{r}
#ped<-read_delim("../Pangolin.incl_HK1.Panel1_2.96.geno_LCWG3.ped",delim="\t",col_names = F)
ped<-read_delim("../Pangolin.inclHK1_8.FinalPanel.geno_LCWG.2mind.rm_dup.ped",delim=" ",col_names = F)

ped %>% filter(X1=="Confiscated") %>% write.table("Pangolin.HK1_8.FinalPanel.94.647ind_genotyped.ped",row.names = F,quote=F,col.names = F,sep="\t")


mix<-ped %>% filter(X1=="Confiscated") %>% dplyr::select(-X3,-X4,-X5,-X6) %>% mutate(sample_type="mixture",repunit="NA") %>% dplyr::select(sample_type,repunit,everything())
str(mix)
mix[,-(1:4)][mix[,-(1:4)] == 0] <- NA
mix[,-(1:4)][mix[,-(1:4)] == "A"] <- "1"
mix[,-(1:4)][mix[,-(1:4)] == "C"] <- "2"
mix[,-(1:4)][mix[,-(1:4)] == "G"] <- "3"
mix[,-(1:4)][mix[,-(1:4)] == "T"] <- "4"

mix %>% write.table("Pangolin.incl_HK1_8.FinalPanel.94.2mind.4rubias.nohead",row.names = F,col.names = F,quote=F,sep="\t")

ped %>% filter(X1!="Confiscated") %>% write.table("Pangolin.reference.FinalPanel.94.geno_LCWG3.ped",row.names = F,quote=F,col.names = F,sep="\t")

ref<-ped %>% filter(X1!="Confiscated") %>% dplyr::select(-X3,-X4,-X5,-X6) %>% mutate(sample_type="reference",repunit=X1) %>% dplyr::select(sample_type,repunit,everything())
str(ref)
ref[,-(1:4)][ref[,-(1:4)] == 0] <- NA
ref[,-(1:4)][ref[,-(1:4)] == "A"] <- "1"
ref[,-(1:4)][ref[,-(1:4)] == "C"] <- "2"
ref[,-(1:4)][ref[,-(1:4)] == "G"] <- "3"
ref[,-(1:4)][ref[,-(1:4)] == "T"] <- "4"

ref %>% write.table("Pangolin.reference.FinalPanel.94.geno_LCWG115.rm_dup.4rubias.nohead",row.names = F,col.names = F,quote=F,sep="\t")

```

Read in this data, and double check that it doesn't have an extra column for whatever reason at the end. 

```{r}
#pang_mix<-read_delim("Pangolin.incl_HK1.Panel1_2.96.geno_LCWG3.rubias_input.txt",delim="\t") %>% distinct()
pang_mix<-read_delim("data/rubias_input/Pangolin.incl_HK1_8.FinalPanel.94.2mind.rubias_input.txt",delim="\t") %>% distinct()


pang_mix_final <- pang_mix %>% mutate_if(is.numeric, as.character)

```

Use the code below to see how many confiscated scales indivdiuals we were able to successfully genotype. 
```{r}
dim(pang_mix_final)
```

Read in the reference file, and make sure all fields are correct type. 
```{r}
pang_ref <- read_delim("data/rubias_input/Pangolin.reference.Panel1_2.94.geno_LCWG.111ind..rubias_input.txt",delim="\t") %>% distinct()
#pang_ref <- read_delim("Pangolin.reference.FinalPanel.94.geno_LCWG115.rm_dup.rmoutlier.rubias_input.txt",delim="\t") %>% distinct() %>% arrange(collection)
pang_ref %>% dplyr::select(repunit) %>% distinct()
pang_ref_final <- pang_ref %>% mutate_if(is.numeric, as.character)

```


__Check how many reference indiviuals are there?__

```{r}
dim(pang_ref_final)
```

To perform genetic mixing and assignment tests, we use the _infer_mixture function_ in rubias. Our data consists of pangolin scales collecetd in multiple sites, however, we just denote this as just "mixture" in the collection column. This is how it is run with the default options:
```{r}
mix_estC <- infer_mixture(reference = pang_ref_final, mixture = pang_mix_final, gen_start_col = 5)
```

The result comes back as a list of four tidy data frames:

1. mixing_proportions: the mixing proportions. The column pi holds the estimated mixing proportion for each collection.

2. indiv_posteriors: this holds, for each individual, the posterior means of group membership in each collection. Column PofZ holds those values. Column log_likelihood holds the log of the probability of the individuals genotype given it is from the collection. Also included are n_non_miss_loci and n_miss_loci which are the number of observed loci and the number of missing loci at the individual. A list column missing_loci contains vectors with the indices (and the names) of the loci that are missing in that individual. It also includes a column z_score which can be used to diagnose fish that don’t belong to any samples in the reference data base (see below).

3. mix_prop_traces: MCMC traces of the mixing proportions for each collection. You will use these if you want to make density estimates of the posterior distribution of the mixing proportions or if you want to compute credible intervals.

4. bootstrapped_proportions: This is NULL in the above example, but if we had chosen method = "PB" then this would be a tibble of bootstrap-corrected reporting unit mixing proportions.

What we are most interested in is the individual posteriors. We can look at the number of individuals assigned to different genetic clusters and associate those individuals to the meta data we have on them: location (MAPS site) and timing (Month,Day,Year).
```{r}
options(scipen = 999)
library(gdata)
library(xlsx)


rep_indiv_estsC <- mix_estC$indiv_posteriors %>% group_by(mixture_collection, indiv, repunit) %>% summarise(rep_pofz = sum(PofZ)) %>% 
  dplyr::select(mixture_collection,indiv,repunit,rep_pofz, everything()) %>% distinct()


rep_indiv_estsC %>% write.table("results/rubias_results/Pangolin.HK_Conf1_8_repunit_rep_pofz.repunit5.rmoutlier.txt",row.names=F,quote=F,sep="\t")
rep_indiv_estsC 
```

What you can see is that each individual has 5 rows, associated with probability of assignment to a repunit (CAR_CON_GAB_DRC, CMR_GAB, Manyu, Western Africa: GH_SL, and Nigeria). 


##Puts this in a more readable format
The above format is a little hard to read, so instead, we are going to use the code below to spread the data- make it so each individual is on one row and the probability of assignment is spread across 5 new columns representing the repunit. The first part writes it to a file, but I want you to look at the output too.
```{r}
spread(data=rep_indiv_estsC,repunit,rep_pofz) %>% dplyr::select(indiv,mixture_collection,CAR_CON_GAB_DRC,CMR_GAB,CMR_Manyu,GH_SL,Nigeria) %>% write.table('results/rubias_results/Pangolin.94.HK_Conf1_8_repunit_spread.repunit5.rmoutlier.txt',quote=F,sep="\t",row.names=F)

spread(data=rep_indiv_estsC,repunit,rep_pofz) %>% dplyr::select(indiv,mixture_collection,CAR_CON_GAB_DRC,CMR_GAB,CMR_Manyu,GH_SL,Nigeria)
```
Read in the data, and rather than having probability of assignment, let's first look at the individuals that are assigned to certain genetic clusters with highest probability- so with certainty. To do that, we say if probability of assignment is greater than 80%, let's call that 1, if not, 0, and then we filter only keeping those individuals with certain assignment using "filter(sum=1)".

5 Repunits

First, we will keep the uncertain data: note the filter `sum==1` is commented out
```{r}
#CMR_GAB and does not filter out uncertain
read_delim("results/rubias_results/Pangolin.94.HK_Conf1_8_repunit_spread.repunit5.rmoutlier.txt",delim="\t") %>% 
  mutate(CAR_CON_GAB_DRC=if_else(CAR_CON_GAB_DRC >.8,1,0)) %>% 
  mutate(CMR_GAB=if_else(CMR_GAB > 0.8,1,0)) %>%
  mutate(GH_SL=if_else(GH_SL > 0.8,1,0)) %>% 
  mutate(Nigeria=if_else(Nigeria > 0.8,1,0)) %>% 
  mutate(CMR_Manyu=if_else(CMR_Manyu > 0.8,1,0)) %>%
  mutate(sum=CAR_CON_GAB_DRC+ CMR_GAB + GH_SL + Nigeria + CMR_Manyu) %>% 
  #filter(sum==1) %>% 
  mutate(assignment=if_else(CAR_CON_GAB_DRC==1,"CAR_CON_GAB_DRC",if_else(CMR_GAB==1,"CMR_GAB",if_else(GH_SL==1,"GH_SL",if_else(Nigeria==1,"Nigeria",if_else(CMR_Manyu==1,"CMR_Manyu","notcertain")))))) %>% 
  dplyr::select(indiv,mixture_collection,CAR_CON_GAB_DRC:Nigeria,assignment)

```

Now we focus on the certain assignments 
```{r}
##certain
read_delim("results/rubias_results/Pangolin.94.HK_Conf1_8_repunit_spread.repunit5.rmoutlier.txt",delim="\t") %>% 
  mutate(CAR_CON_GAB_DRC=if_else(CAR_CON_GAB_DRC > 0.8,1,0)) %>% 
  mutate(CMR_GAB=if_else(CMR_GAB > 0.8,1,0)) %>%
  mutate(GH_SL=if_else(GH_SL > 0.8,1,0)) %>% 
  mutate(Nigeria=if_else(Nigeria > 0.8,1,0)) %>% 
  mutate(CMR_Manyu=if_else(CMR_Manyu > 0.8,1,0)) %>%
  mutate(sum=CAR_CON_GAB_DRC+CMR_GAB+GH_SL+Nigeria+CMR_Manyu) %>% 
  filter(sum==1) %>% 
  mutate(assignment=if_else(CAR_CON_GAB_DRC==1,"CAR_CON_GAB_DRC",if_else(CMR_GAB==1,"CMR_GAB",if_else(GH_SL==1,"GH_SL",if_else(Nigeria==1,"Nigeria",if_else(CMR_Manyu==1,"CMR_Manyu","notcertrain")))))) %>% 
  dplyr::select(indiv,mixture_collection,CAR_CON_GAB_DRC:Nigeria,assignment) %>% rename(Sample=indiv) %>%  write.table("results/rubias_results/Pangolin.94.HK_Conf1_8_repunit_withAssignments01.repunit5.rmoutlier.txt",quote=F,sep="\t",row.names=F)

#Tallies the pangolin assignments that are certain

assign01<-read_delim("results/rubias_results/Pangolin.94.HK_Conf1_8_repunit_withAssignments01.repunit5.rmoutlier.txt",delim="\t") 
assign01 %>% group_by(assignment) %>% tally()
```
__How many individuals were assigned with certainty to each distinct genetic cluster?__

__How many individuals were not assigned with certainty? And what was their probaiblity for each genetic cluster?__

By seizure
```{r}
pedM<-read_delim("data/Pangolin.HK1_8.FinalPanel.94.647ind_genotyped.ped",delim="\t",col_names = F) %>% rename(Sample=X2)

meta_allM<-read.xlsx("data/Pangolin_Sample_Tracker_v20_Feb152022.xlsx",sheetIndex =2) %>% rename(Sample=UCLA.sample.ID)

meta_allM2<-pedM %>% left_join(meta_allM) %>% dplyr::select(-c(X3:X194))
 `%notin%`<-Negate(`%in%`)
missing<-meta_allM %>% filter(Sample %notin% pedM$Sample)

assign01 %>% left_join(meta_allM2) %>% group_by(	
Seizure.Number,	
Declared.country.of.origin,assignment) %>% tally() %>% arrange(Declared.country.of.origin) %>% write.table("results/rubias_results/Pangolin.94.HK_Conf1_8_repunit5_withAssignments01.bySeizure.txt",row.names=F,quote=F,sep="\t")
```

Use the code below to answer this question: Where are the uncertain assignments?
```{r}
`%notin%` <- Negate(`%in%`)

uncertain<-read_delim("results/rubias_results/Pangolin.94.HK_Conf1_8_repunit_spread.repunit5.rmoutlier.txt",delim="\t") %>% filter(indiv %notin% assign01$Sample)
uncertain
```

Self assessment of breeders (georeferenced pangolins) to account for ascertainment bias- so excluding the LCWG individuals. 


```{r}
`%notin%`<-negate(`%in%`)

t<-read.xlsx("data/rubias_input/Pangolin_tests_assertainment.xlsx",sheetIndex = 1) 

pang_ref2<-read_delim("data/rubias_input/Pangolin.reference.Panel1_2.94.geno_only.rmoutlier.rubias_input.newname.txt",delim="\t")

pang_ref_final2 <- pang_ref2 %>% filter(indiv %notin% t$ind_removed) %>% mutate_if(is.numeric, as.character)
dim(pang_ref_final2)

breed_sa2 <- self_assign(pang_ref_final2, 5)
```

Formatting table
```{r, rows.print = 40}
ind<-read_delim("data/rubias_input/Pangolin.reference.Panel1_2.94.geno_only.rmoutlier.rubias_input.newname.txt",delim="\t") %>% dplyr::select(indiv)
dim(ind)
#old sample file: /Pangolin_sample_trackerv17.08182021.xlsx
breeding_meta2<-read.xlsx("data/Pangolin_sample_trackerv17.08182021.xlsx",sheetIndex=1) %>% filter(
UCLA.sample.ID %in% ind$indiv)


sa2<-breeding_meta2 %>%
  dplyr::select(Location, Sample_name2) %>%
  left_join(breed_sa2, ., by = c("indiv" = "Sample_name2")) %>%
  group_by(Location, indiv) %>%
  slice(1) %>%
  ungroup() %>%
  count(Location, inferred_repunit) %>%
  pivot_wider(
    names_from = inferred_repunit,
    values_from = n,
    values_fill = list(n = 0L)
  ) %>% 
  dplyr::select(
    Location,
    CAR_CON_GAB_DRC,
    CMR_GAB,
    CMR_Manyu,
    Nigeria,
    GH_SL
  )

sa2 %>% write.table("results/rubias_results/Pang.94.self_assessment.rm_ind_4ascertain.rmoutlier.genotype_only.txt",row.names = F,quote=F,sep="\t")
sa2

breeding_meta2 %>%
  dplyr::select(Location, Sample_name) %>%
  left_join(breed_sa2, ., by = c("indiv" = "Sample_name")) %>%
  group_by(Location, indiv) %>%
  slice(1)  %>% dplyr::select(-missing_loci) %>% 
  write.table("results/rubias_results/Pangolin.self_assessment.geno_only.rubias.InferGroup.txt",row.names = F,quote=F,sep="\t")

```

#### Breeding Pangolin Shapefile is needed for OriGen

Finally, we need to have a GIS Shapefile that tells us the range of the white-bellied pangolins, so that genoscape can be clipped properly.  We read this shapefile with the 
`st_read()` function from package `sf`. 
See code below. This also comes in handy for looking at probabilty of spatial assignment in Origen, the second analysis we'll do today.

```{r}
breeding <- st_read("data/Phataginus_tricuspis_shapefile/data_0.shp")

```

##Run Origen to get predicted Lat/Long of each individual. 
While rubias assigns to a distinct genetic cluster, we might want to know actual locations. Moreover, having the lat/long of each individual allows us to include those intermediate individuals as those might have intermediate locations.

Here you have to input 5 data files. The ped and map file for breeders, with an associated location file (Sample, lat and long), and the ped and map file for the unknown confiscated pangolin scales

```{r}

pangOri<-ConvertUnknownPEDData("data/OriGen_input//Pangolin.breeders.FinalPanel.geno_LCWG.2mind.rm_dup.name_fix.rm3outliers","data/OriGen_input/Pangolin.111breeding_b.4Origen.loc","data/OriGen_input/Pangolin.HK1_8.FinalPanel.94.647ind_genotyped")


#Read in breeding locations
breed_loc<-read_delim("data/OriGen_input/Pangolin.111breeding_b.4Origen.loc",delim="\t")

breed_meta<-read.xlsx("data/Pangolin_Sample_Tracker_v20_Feb152022.xlsx",sheetIndex = 1) %>% rename(Sample=UCLA.sample.ID) %>% dplyr::select(-Longitude,-Latitude)
breed_meta2<-breed_loc %>% left_join(breed_meta)
breed_meta2
```

To run Origen, we are going to use the FitOriGenModelFindUnknowns model. You can play around with the MaxGridLength to adjust the size of pixels/grids you can pinpoint the breeding origin locaiton of the confiscated scales.
```{r}
#MaxGridLength is the maximum number of boxes allowed to span the region in either direction
#RhoParameter is a tuning constant

OrigenP<-FitOriGenModelFindUnknowns(pangOri$DataArray,pangOri$SampleCoordinates,pangOri$UnknownData,MaxGridLength=70,RhoParameter=10)

```


Now, we are  taking the white-bellied pangolin shapefile, and we are only going to keep the lat/longs that have the highest probability if they fall within the breeding range They should normally, and in many cases you'll see them fall near where we have sampling localities, but this makes sure this is the case.
```{r}
#convert the geometry of the sf to spatial polygons
breedE_spd1<-sf::as_Spatial(st_geometry(breeding),IDs=as.character(1:nrow(breeding)))

#grab data from sf object
df<-breeding
df$geometry<-NULL
df<-as.data.frame(df)

#create teh spatialpolygon data fram
breedE_spd<-sp::SpatialPolygonsDataFrame(breedE_spd1,data=df)


#now try to subset grid
grid <- expand.grid(OrigenP$GridCoordinates[1,],OrigenP$GridCoordinates[2,])
grid <- grid[grid$Var2!=0,]
names(grid) <- c("x","y")
pts <- SpatialPointsDataFrame(grid[,c(1,2)],data=grid[,c(2,1)],proj4string=attributes(breedE_spd)$proj4string)
overlap <- sp::over(pts,as(breedE_spd,"SpatialPolygons"))
subgrid <- grid[!is.na(overlap),]
subgrid$xind <- match(subgrid$x,OrigenP$GridCoordinates[1,])
subgrid$yind <- match(subgrid$y,OrigenP$GridCoordinates[2,])

subgrid

for (i in 1:nrow(OrigenP$UnknownData)) {
  myGrid2 <- OrigenP$UnknownGrids[,,i]
  for (j in 1:nrow(subgrid)) {
    subgrid[j,i+4] <- myGrid2[subgrid$xind[j],subgrid$yind[j]]
  }
}
```

This aligns the the data with the confiscated pangolin sample names, and writes out the data so you don't have to run this analysis each time. If you ever need to come back to the data, just read in the RDS file.
```{r}
#POPID,  K=5,
pedM<-read_delim("data/OriGen_input/Pangolin.HK1_8.FinalPanel.94.647ind_genotyped.ped",delim="\t",col_names = F) %>% rename(Sample=X2)


names(subgrid) <- c("x","y","xind","yind",as.character(pedM$Sample))
saveRDS(subgrid,"results/OriGen_results/Pang_OriGen_ProbSurface.HK1_8.POPID.rds")

rds<-readRDS("results/OriGen_results/Pang_OriGen_ProbSurface.HK1_8.POPID.rds")
rds
```

Each row is associated with a lat/long and each indivdiual has a probability of being assigned to that lat/long. In this code chunk, we are chossing the lat/long with the highest probability.
```{r}
write.table(rds, "results/OriGen_results/Pang_OriGen_ProbSurface.HK1_8.POPID.txt", quote = F, sep="\t", row.names = F)

df_rds<-read_delim("results/OriGen_results/Pang_OriGen_ProbSurface.HK1_8.POPID.txt",delim="\t") 

#column where individual starts
for (j in 5:ncol(df_rds))
{
  i <- max(df_rds[,j],na.rm=T) 
  k <- df_rds[which(df_rds[,j] == i), c(1,2,j)] 
  write.table(k,file="results/OriGen_results/Pang_OriGen_ProbSurface.94.HK1_8.POPID.highestprob_origen.final.txt",row.names=F,quote=F,sep="\t",append=T,col.names=F)
}

##read in the file you just created. 
pred<-read_delim("results/OriGen_results/Pang_OriGen_ProbSurface.94.HK1_8.POPID.highestprob_origen.final.txt",delim="\t",col_names = F) %>% rename(predLong=X1,predLat=X2,highprob=X3)

meta_latlong<-read.xlsx("data/Pangolin_Sample_Tracker_v20_Feb152022.xlsx",sheetIndex = 2) %>% rename(Sample=UCLA.sample.ID)
ped<-read_delim("data/OriGen_input/Pangolin.HK1_8.FinalPanel.94.647ind_genotyped.ped",col_names = F) %>% rename(Sample=X2) %>% dplyr::select(-c(X3:X194))
conf_latlong<-ped %>% left_join(meta_latlong)
dim(pred)
##There are no sample names attached to it, BUT it is in the order of the meta data for the confiscated pangolins (aka the order of the ped file), so we are adding those sample names to the file and writing out the data.
out<-cbind(conf_latlong,pred) %>% dplyr::select(-c(`NA.`:`NA..20`))
saveRDS(out,"results/OriGen_results/Pang_OriGen_ProbSurface.94.HK1_8.POPID.predlatlong_only.clipped.w_meta.rds")
write.table(out,"results/OriGen_results/Pang_OriGen_ProbSurface.94.HK1_8.POPID.predlatlong_only.clipped.w_meta.txt",row.names = F,quote=F,sep="\t")

```
Let's look at seizures
```{r}
out %>% group_by(Seizure.Number,Declared.country.of.origin) %>% tally()

```


Let's plot the probabilities of each lat/long location. We can see which ones have extremely low probabilities and potentially filter out those.
```{r}
g<-gghistogram(out, x = "highprob",
            add = "mean", rug = TRUE,
            palette = "Dark2",bins=60)+theme(aspect.ratio = 1)


pdf("results/OriGen_results/Pang.highprob_Origen.hist.pdf")
g
dev.off()
library(RColorBrewer)
out %>% filter(highprob>0.1) %>% gghistogram(x = "highprob",
            add = "mean", rug = TRUE,
            palette = "Dark2",bins=60)+theme(aspect.ratio = 1)
```

